Developing Observational Health Research Capacity in the Julia Ecosystem

### 2023-12-14 paper.tex
Added more todos and restructuring

% EXPLAIN: Why is this a crucial innovation?
% EXPLAIN: What is meant by a transferable analysis
% TODO: Cite: J. M. Overhage, P. B. Ryan, C. G. Reich, A. G. Hartzema, and P. E. Stang, “Validation of a common data model for active safety surveillance research,” Journal of the American Medical Informatics Association, vol. 19, no. 1, pp. 54–60, 2012.
Arguably, the most crucial innovation that has emerged is the Observational Medical Outcomes Partnership Common Data Model (OMOP CDM) which was designed specifically to standardize observational health data to enable transferable analyses.
% TODO: Add citation
% NOTE: Why is mapping relevant here?
The importance of this model has become apparent within open science organizations like the OHDSI (Observational Health Data Science and Informatics) network where members across $100+$ countries with access to various health systems have been able to successfully map their patient data to the OMOP CDM.
% TODO: Add citation about successful collaborations
% EXPLAIN: What is a network study?
% TODO: Cite the quote "write once, run everywhere" 
% http://www.sun.com/smi/Press/sunflash/1996-01/sunflash.960123.10561.xml} 
This has led to successful collaboration relationships across countries where one research group can develop a network study and then disseminate the study across collaborators leading to the idea that you can "write once, run everywhere."

% TODO: Use this to address the notion of what "network studies" are
In particular, one collaboration mechanism that OHDSI has been able to operationalize meaningfully while working with observational health data is the notion of "network studies".
These studies, are designed to be transparent and encompasses research endeavors being explored by members of the OHDSI comprehensive of relevant documentation, analysis code, results, and study protocols describing the study's scope and objectives.
One core innovation brought about by OHDSI in the space of network science is how, given the HADES tools and procedures established by OHDSI, a research group can develop their analysis on their own OMOP CDM instance and then bundle together their study.
Once this packaging is done, they can pass along the study package to another site to collaborate on the same analysis but on different data assets.
Data confidentiality is preserved in this approach as data from individual participating sites remains exclusive to each site, with only aggregate results shared across a given collaborator network network.
As such, no risk for re-identification comes about through this network collaboration and has ushered in the possibility for similar research approaches to be taken across the globe. \cite{ohdsi2019book}

\subsection{Background}

\subsubsection{Core Technologies within Observational Health Research}

% TODO: Reframe this paragraph to make it like a better background introduction.
What could be considered the magnum opus of the OMOP group was the development of what is called the OMOP Common Data Model (OMOP CDM) to house and ingest this sort of data in a consistent format that has been optimized to assist in extracting meaningful population health information.
In brief, the OMOP Common Data Model (CDM) is a "person-centered" model that has standardized common events in a given patient encounter with database tables that represent the unique structures found in patient events.
This creates a longitudinal view of all healthcare-related events per person recorded in a singular health system.
As the OMOP CDM has become a cornerstone in running these widely distributed network studies, special attention to developing tools around handling OMOP CDM data is necessary to meaningfully contribute within the observational health research space.
The OMOP CDM itself does not require a specific technology to work with the data stored in this standard.
It features a person-centric design where each domain records personal identity while prioritizing data protection through the limiting of information that could endanger patient anonymity.

\subsubsection{The Impact of the Observational Health Data Science and Informatics (OHDSI) Consortium}

% TODO: Reframe this to suggest the more general impact that OHDSI has had within the observational health research space.
Driving development within the subecosystem is the influence the OHDSI open-science network.
As mentioned previously, OHDSI is a global, collaborative, interdisciplinary, and open network of researchers, healthcare professionals, students, patients, and other entities committed to large-scale health data analytics and generating reliable evidence, with the goal of promoting better health decisions and better care. 
OHDSI had its inception in the development of the Observational Medical Outcomes Partnership (OMOP) that was created in 2009 by the US Food and Drug Administration to address concerns related to data type, study design, and data privacy concerns of "Real World Data" \cite{ohdsi2019book} \cite{FDARealWorldEvidence}
OHDSI was developed to focus on leveraging observational health data stored within the OMOP CDM.
Since current observational health data sources are aggregated at different levels and cannot be generalized to the overall population, the data is often incomplete, inaccurate, and prone to issues of bias, error handling, and multidimensional complexity. 
The goal of OHDSI is to produce reproducible, robust, and reliable evidence by developing standardized data formats and research protocol designs to enable large-scale collaborative research and improve the transparency and scalability of observational health research.

OHDSI supports the development of open-source software tools in a centralized ecosystem called HADES (Health Analytics Data-to-Evidence Suite) for particularly managing and working with OMOP CDM formatted data.
The HADES ecosystem has been in existence for $\approx 10$ years and has given rise to improvements in the quality and reliability of observational health research as well as motivating novel collaboration mechanisms.

\subsubsection{JuliaHealth's Role within Observational Health Research}

@@

\section{Describing Common Observational Health Research Tasks within the Julia Ecosystem}

>>>>

% EXPLAIN: Why is this a crucial innovation?
% EXPLAIN: What is meant by a transferable analysis
% TODO: Cite: J. M. Overhage, P. B. Ryan, C. G. Reich, A. G. Hartzema, and P. E. Stang, “Validation of a common data model for active safety surveillance research,” Journal of the American Medical Informatics Association, vol. 19, no. 1, pp. 54–60, 2012.
Arguably, the most crucial innovation that has emerged is the Observational Medical Outcomes Partnership Common Data Model (OMOP CDM) which was designed specifically to standardize observational health data to enable transferable analyses.
% TODO: Add citation from OHDSI website's collaborator statistics
% EXPLAIN: Why is mapping relevant here?
The importance of this model has become apparent within open science organizations like the OHDSI (Observational Health Data Science and Informatics) network where members across $100+$ countries with access to various health systems have been able to successfully map their patient data to the OMOP CDM.
% TODO: Add citation about successful collaborations
% EXPLAIN: What is a network study?
% TODO: Cite the quote "write once, run everywhere" 
% http://www.sun.com/smi/Press/sunflash/1996-01/sunflash.960123.10561.xml} 
This has led to successful collaboration relationships across countries where one research group can develop a network study and then disseminate the study across collaborators leading to the idea that you can "write once, run everywhere."

\subsection{Core Technologies}

Modern observational health research centers around a variety of core technologies.
A few of these technologies highlighted in this paper are:

\item Data standards within observational health research
\item Collaborative networks and studies using observational health data
\item Research software to support the analysis of observational health data

In particular, the data standard that will be most highlighted for the remainder of this paper is the Observational Medical Outcomes Partnership Common Data Model (OMOP CDM).
Subsequent discussion will be focused on the opportunities that this specific data model has given rise to across observational health as well as research communities such as the Observational Health Data Sciences and Informatics (OHDSI) open science consortium.

% TODO: This paragraph should be centered on the core technology of data standards -- in particular, the OMOP CDM
% EXPLAIN: What is real world data? Use https://jacobzelko.com/10282021140730-real-world-evidence/
OMOP was created in 2009 by the US Food and Drug Administration to address concerns related to data type, study design, and data privacy concerns of "Real World Data" \cite{ohdsi2019book} \cite{FDARealWorldEvidence}
What could arguably be considered the longest standing achievement of OMOP was the development of the OMOP Common Data Model (OMOP CDM) to house and ingest real world data.
% TODO: Combine the next two sentences together
% EXPLAIN: What does "person-centered" mean?
% EXPLAIN: What is a patient event? Could you give an example?
in a consistent format that has been optimized to assist in extracting meaningful population health information.
In brief, the OMOP Common Data Model (CDM) is a "person-centered" model that has standardized common events in a given patient encounter with database tables that represent the unique structures found in patient events.
% TODO: Develop this fragment into a full sentence
prioritizing data protection through the limiting of information that could endanger patient anonymity.
% EXPLAIN: What is a health system? Could you give an example?
This creates a longitudinal view of all healthcare-related events per person recorded in a singular health system.
% TODO: Make this more coherent to the rest of the paragraph
% TODO: Explicitly mention that although the model prescribes a database schema, it does not need a specific database type
The OMOP CDM itself does not require a specific technology to work with the data stored in this standard.

% TODO: This paragraph should be centered on the core technology of the concept of network studies
% TODO: Use this to address the notion of what "network studies" are
% TODO: Make this initially less OHDSI specific; pull from https://jacobzelko.com/05122023034151-network-macarthur
In particular, one collaboration mechanism that OHDSI has been able to operationalize meaningfully while working with observational health data is the notion of "network studies".
These studies, are designed to be transparent and encompasses research endeavors being explored by members of OHDSI comprehensive of relevant documentation, analysis code, results, and study protocols describing the study's scope and objectives.
OHDSI had its inception in the development of the Observational Medical Outcomes Partnership (OMOP) that was created in 2009 by the US Food and Drug Administration to address concerns related to data type, study design, and data privacy concerns of "Real World Data" \cite{ohdsi2019book} \cite{FDARealWorldEvidence}
OHDSI was developed to focus on leveraging observational health data stored within the OMOP CDM.
Since current observational health data sources are aggregated at different levels and cannot be generalized to the overall population, the data is often incomplete, inaccurate, and prone to issues of bias, error handling, and multidimensional complexity. 
The goal of OHDSI is to produce reproducible, robust, and reliable evidence by developing standardized data formats and research protocol designs to enable large-scale collaborative research and improve the transparency and scalability of observational health research. 
% TODO: Carry HADES discussion into the paragraph talking about research tooling
One core innovation brought about by OHDSI in the space of network science is how, given the HADES tools and procedures established by OHDSI, 
% TODO: Make this more general such that it flows into the next paragraph on research tooling better
a research group can develop their analysis on their own OMOP CDM instance and then bundle together their study.
% EXPLAIN: What is a study package?
% EXPLAIN: What data assets?
Once this packaging is done, they can pass along the study package to another site to collaborate on the same analysis but on different data assets.
% EXPLAIN: What is a collaborator network?
Data confidentiality is preserved in this approach as data from individual participating sites remains exclusive to each site, with only aggregate results shared across a given collaborator network.
As such, no risk for re-identification comes about through this network collaboration and has ushered in the possibility for similar research approaches to be taken across the globe. \cite{ohdsi2019book}

% TODO: Develop this paragraph more about research software
As the OMOP CDM has become a cornerstone in running these widely distributed network studies, special attention to developing tools around handling OMOP CDM data is necessary to meaningfully contribute within the observational health research space.
OHDSI supports the development of open-source software tools in a centralized ecosystem called HADES (Health Analytics Data-to-Evidence Suite) for particularly managing and working with OMOP CDM formatted data.
The HADES ecosystem has been in existence for $\approx 10$ years and has given rise to improvements in the quality and reliability of observational health research as well as motivating novel collaboration mechanisms.

\subsection{JuliaHealth and Observational Health Research}

@@

\section{Accomplishing Common Observational Health Research Tasks within Julia Ecosystem}
